---
title: "Tbll_xtabs"
author: "Wolfgang Peter"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  md_document:
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{Tbll_xtabs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(stp25output2)
require(stp25stat2)
require(stp25tools)
require(stp25plot)
library(vcd)
```

## Mosaic-Plot

Ein Mosaikplot ist eine grafische Repräsentation einer
Kontingenztabelle, in der die Fläche der Rechtecke des Plots
proportional zu den Häufigkeiten der Kontingenztabelle ist.

A mosaic plot is a graphical display of the cell frequencies of a
contingency table in which the area of boxes of the plot are
proportional to the cell frequencies of the contingency table.

```{r, include=FALSE}
dat<- 
stp25aggregate::GetData("
   sex  racing  diet Freq
 female   10km  omni  147
   male   10km  omni   76
 female     HM  omni  238
   male     HM  omni  197
 female   M/UM  omni  215
   male   M/UM  omni  399
 female   10km  vegi  113
   male   10km  vegi   29
 female     HM  vegi  143
   male     HM  vegi   72
 female   M/UM  vegi  125
   male   M/UM  vegi  116
 female   10km  vegan  208
   male   10km  vegan   49
 female     HM  vegan  271
   male     HM  vegan  111
 female   M/UM  vegan  168
   male   M/UM  vegan  187                       
                        ")
# xtabs(Freq~ sex, dat)



```

```{r, results='asis'}
x1<- xtabs(Freq~ diet+ sex, dat)
Tbll_xtabs( xtabs(Freq~ diet +sex, dat), margin = "sex" ) %>% 
  Output()

```

Table `r Tab_Index` shows data on the relation between sex and diet
among `r sum(dat$Freq)` runners.

The Pearson chi-Squared for this data is `r APA(summary(x1))`,
indicating substantial departure from independence.

Figure 1 shows the expected frequencies for sex and diet. The heights of
the boxes are proportional to the percentage of females and males,
`r round(prop.table(xtabs(Freq~ sex, dat))[1]*100)` % of the runners
were female. The widths of the boxes are proportional to the percentage
of diet.

There are more male omnivorous runners and female vegan runners that
would occur under independence, and fewer female omnivorous runners and
male vegan runners.

```{r, fig.cap='Mosaic-Plot displaying the expected frequencies. Color and Shading showing the standardized deviation from independenze by the color shading of each rectangle. Cells with positive deviation are blue, negative deviation are drawn red. The absolut value of the deviation is portrayed by schading density.',fig.height=4, fig.width=4}

ipol <-function(x) pmin(x/6,1)
mosaic(Freq
  ~ diet | sex,
  dat,
  shade = TRUE,
  legend = TRUE ,
  #labeling=labeling_left,
  varnames = F, 
  gp=shading_hcl, gp_args=list(interpolate=ipol),
  # ?labeling_border
  labeling_args = list(abbreviate_labs = c(8, 5, 4)),
   main ="Tow-Way Mosaic"
)

```

```{r}
vcd::assocstats(x1) %>% Tbll()

glm(Freq
    ~ diet * sex,
    dat, family = poisson() ) %>% car::Anova()

```

```{r, fig.height=7, fig.width=7}
ipol <-function(x) pmin(x/6,1)
mosaic(Freq
  ~ racing + diet | sex,
  dat,
  shade = TRUE,
  legend = TRUE ,
  #labeling=labeling_left,
  varnames = F, 
  gp=shading_hcl, gp_args=list(interpolate=ipol),
  # ?labeling_border
  labeling_args = list(abbreviate_labs = c(8, 5, 4)))

```

## Loglinear Models

Kontingenztabellen können auch mit log-linearen Modellen untersucht
werden.

Im log-lin-Modellen wird die Abhängigkeit (assoziation) zwischen den
Faktoren untersucht. Es wird aber nicht zwischen **Ziel** und
**Einflussvariable** unterschieden. Die Fallzahl soll das 5-fache der
betrachteten Zellen betragen. Bei dreidimensionalen Tafeln sind 5
verschiedene Modelle zu testen.

| Model                    | Model formula      | Symbol       |
|--------------------------|--------------------|--------------|
| Mutual independence      | \~A + B + C        | [A][B][C]    |
| Joint independence       | \~A\*B + C         | [AB][C]      |
| Conditional independence | \~(A+B)\*C         | [AC][BC]     |
| All two-way associations | \~A*B + A*C + B\*C | [AB][AC][BC] |
| Saturated model          | \~A\**B\**C        | [ABC]        |

### Beispiel Marital Status and Permarital and Extramarital Sex

    G Gender
    M Familienstand 
    P voreheliches Sex
    E außereheliches Sex

```{r, echo=FALSE, results='asis'}
require(stpvers)
require(vcd)

library(MASS)

dat2<- 
  stp25aggregate::GetData("
gender  premarital extramarital marital Freq
 male n n  divorced 68
 male n y divorced 17
 male y n divorced 60
 male y y divorced 28
 female n n   divorced 214
 female n y divorced 36
 female y n divorced 54
 female y y divorced 17
 male n n  marrid 130
 male n y marrid 4
 male y n marrid 42
 male y y marrid 11
 female n n   marrid 322
 female n y marrid 4
 female y n marrid 25
 female y y marrid 4
                        ", output=FALSE)

x1<-xtabs(Freq ~ gender + premarital + extramarital + marital, dat2)
x.g.p<-xtabs(Freq ~ gender + premarital  , dat2)
x.g.p.e.m <- xtabs(Freq ~ gender + premarital + extramarital + marital, dat2)
xtabs(Freq ~ gender + premarital + extramarital + marital, dat2) %>% 
  ftable() %>%
  stp25tools::fix_to_df() %>% Output("Familienstand")

```

```{r}
mart_tab1 <- xtabs(Freq ~ gender + premarital  , dat2)
(mart.1 <- MASS::loglm(~ gender + premarital, data = mart_tab1))
(OR <- oddsratio(mart_tab1, log = FALSE)) # odds ratio
```

```{r, include=FALSE}
mart_tab2 <- xtabs(Freq ~ gender + premarital + extramarital , dat2)

oddsratio(xtabs(Freq ~   premarital + extramarital +gender, dat2), log=FALSE)

mart_tab3 <-
  xtabs(Freq ~ gender + premarital + extramarital + marital , dat2)


```

```{r}
mart.1 <- 
  loglm( ~ gender + premarital, 
          data = mart_tab1)
mart.2 <-
  loglm(~ gender * premarital + extramarital,
        data = mart_tab2)

mart.3 <-
  loglm(~ gender * premarital * extramarital + marital, 
        data = mart_tab3)

mart.4 <-
  loglm(~ gender + premarital + extramarital + marital, 
        data = mart_tab3)

```

```{r, , include=FALSE, results='asis'}
t1<- mart.1 %>% Tbll() 
t2<-mart.2 %>% Tbll()
t3<-mart.3 %>% Tbll()
t4<-mart.4 %>% Tbll()
 

rslt<-rbind( t1[1,], t2[1,], t3[1,], t4[1,])
rslt$Test <- c("[G][P]","[GP][E]","[GPE][M]","[G][P][E][M]")
rslt %>% Output("Likelihood Ratio G²")
 
```

Die Untersuchung der Scheidungsrate erfolgt in meheren Schritten. Als
erstes untersuchen wir ist voreheliches Sex vom Gechlecht abhängig.
[G][P]: P as response to G

Als zweites wird untersucht ob das Geschlecht und voreheliches
Sexzusammen einen Effect auf außereheliches Sex haben [GP][E] E as
response to G and P jointly.

Und in den Weiteren Schritten wird der Familienstand in die Analyse
eingebracht.

Das [G][P] Model zeigt einen hohen Zusammenhang von G und P mit G² =
`r t1[1,2]`

     For the model  G² = `r t1[1,2]` indicating that Gender and Premarital sex are highly assocated. The mosaic plot shows that mean are much more likely to report Premarital sex than are woman.

```{r, echo=FALSE}
 plot(mart.1, main="[G][P]")  
  plot(mart.2, main="[GP][E]")
 
 
plot( mart.3, main="[GPE][M]")
plot( mart.4, main="[G][P][E][M]")
```

## Zweifach - Tabelle N x M

```{r}
data(infert, package = "datasets")
infert$case  <- factor(infert$case ,1:0, c("case", "control") )
head(infert)
```

```{r usage-2, results='asis'}

Tbll_xtabs( ~ education + case, infert, include.percent = FALSE) %>%
  Output()

Tbll_xtabs( ~ education + case,
            infert,
            include.total = TRUE,
            include.count = FALSE) %>%
  Output()

Tbll_xtabs(
  ~ education + case,
  infert,
  include.total.columns = TRUE,
  include.count = FALSE
) %>%
  Output()

Tbll_xtabs( ~ education + case,
            infert,
            include.total.rows = TRUE,
            include.count = FALSE) %>%
  Output()
 
 

```

## Dreifach - Tabelle N x M x O

Feinkontrolle über die Berechnung und Darstellung der Prozent.

:   education + induced + case

-   N und M Zeilen
-   O Spalten (Letztes Item ist die Spalte)

Fall 1 Spalten - Prozent case und control je 100%

```{r}
x <- xtabs( ~ education + induced + case, infert)

#  margin = 3 ist equivalent mit  margin = "case"
#  addmargins(x, c(2,3)) == addmargins(x, c(3,2)) 
round(
  addmargins(
    prop.table(x, margin = "case"), 
    c(1, 2)
    ), 2) * 100

```

-   margin geht an *prop.table* (1 indicates rows, 2 indicates columns)
-   add.margins geht an *addmargins* mit margin=seq_along(dim(x))

```{r usage-margin, results='asis'}
Tbll_xtabs(
  x,
  margin = "case",
  add.margins = c("education", "induced"),
  include.count = FALSE) %>% 
  Output()

Tbll_xtabs(
  x,
  margin = "case",
 # add.margins = c("education", "induced"),
  include.count = FALSE) %>% 
  Output()

```

```{r data, results='asis'}
dat <- Label(
  GetData(
"sex treatment  neg  pos
f   KG         10	   9
f   UG         14	   5
m   KG         23	   7
m   UG         18	  14",
Tabel_Expand = TRUE,
id.vars = 1:2,
value = "befund"
  ), sex = "Gexhlecht", treatment = "Behandlung")

dat %>% Tabelle2(sex, treatment, befund)

```

    APA2(xtabs( ~ sex +treatment + befund, dat)) 



    APA2(xtabs( ~ sex +treatment + befund, dat),
         include.total = TRUE))

Mehr als kalter Kaffee

```{r simple2-tabs, results='asis'}

Tbll_xtabs(xtabs( ~ sex +treatment + befund, dat),
     include.total.rows = TRUE,
     include.total.columns = TRUE) %>%   
  Output("Einfache Tabelle")


Tbll_xtabs(xtabs( ~ sex +treatment + befund, dat),
     include.total.columns = FALSE) %>%   
  Output("Einfache Tabelle")

Tbll_xtabs(xtabs( ~ sex +treatment + befund, dat),
     include.total.columns = TRUE) %>%   
  Output("Einfache Tabelle")

```

## Library vcd

```{r}
 
x1 <- xtabs(~ sex + treatment, dat)
vcd::assocstats(x1)

Tbll_xtabs(x1, include.test = TRUE)[[2]]
#Tbll_xtabs(x1, include.prop.chisq = TRUE)[[2]] #Proportion noch nicht fertig.
Tbll_xtabs(x1, include.chisq = TRUE)[[2]]
Tbll_xtabs(x1, include.correlation = TRUE)[[2]] 



 
```

```{r}

x2 <- xtabs( ~ sex +treatment + befund, dat) 
vcd::assocstats(x2)
```
